Correcting for sample selection in Bayesian analyses

### 2017-06-08 papers/general_method/general.tex
Updating paper
Adding github links

 In this section we present sample selections of increasing complexity. Each sample selection is accompanied by an illustrative example inspired by Type Ia supernova cosmology, where we characterize the properties of a standard or standardizable candle.

@@

We thus begin by considering an ideal situation where the sample is complete.  All events are included in the sample such that $P(S | {\rm data}, \theta)=1$.  Trivially this expression is independent of our data and model parameters $\theta$, and our likelihood from Equation \eqref{eq:main} reduces down to Equation \eqref{eq:like}. As a concrete example of this case, let us consider a model for a population of objects whose brightnesses form a normal distribution with average $\mu$ and standard deviation $\sigma$. Let us also assume that our experiment produces data $x_i$, which are independent measurements of unique objects' brightnesses (with negligible measurement uncertainty), which we can formally state as
\begin{align}
\vec{x} \sim \mathcal{N}(\mu,\sigma).
\end{align}
If, having collected our observations $\vec{x}$, we wanted to constrain $\mu$ and $\sigma$, this would be a simple task of modelling the posterior surface. We simply wish to map the surface
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
where our model parameters $\theta = \lbrace \mu, \sigma \rbrace$ and our data is given by $\vec{x}$.
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto P(\vec{x} | \mu, \sigma) P(\mu, \sigma)
\end{align}
With uniform priors, $P(\mu,\sigma) = {\rm constant}$, and can be absorbed into the constant of proportionality. Expanding our observation vector to take into account all observations and not just one, the posterior surface is given by
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto \prod_{i=1}^N \mathcal{N}(x_i | \mu, \sigma), \label{eq:prod}
\end{align}
where $\mathcal{N}(x_i|\mu,\sigma)$ represents a the probability of drawing sample $x_i$ from the Normal distribution with mean $\mu$ and standard deviation $\sigma$. Generating a hundred data points with $\mu=100,\ \sigma=10$, we can recover our input parameters easily, as shown in Figure \ref{fig:perfect}.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/perfect.pdf}
	\end{center}
	\caption{A systematic test of our perfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points. Any systematic offset in our model would be revealed by a shift in the stacked results away from the true parameter values.}
	\label{fig:perfect}
\end{figure}

@@

With this sample selection, all events satisfy $x_i > \alpha$, giving $P(S|x,\theta) = \mathcal{H}(x - \alpha)$, where $\mathcal{H}$ is the Heaviside step function
\begin{align}
\mathcal{H}(y)\equiv \begin{cases}
1 \quad {\rm if }\  y \ge 0 \\
0 \quad {\rm otherwise.}
\end{cases}
\end{align} 
We assign an arbitrary value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates. However, we can correct for this truncation using Equation \eqref{eq:main}, as the integral in the denominator has an analytic solution. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in Equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is independent of our observations $\vec{x}$, we can take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma),
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates. This is demonstrated in Figure \ref{fig:imperfect}, which shows the posterior surfaces for when you take sample selection into account and when you do not. The selection bias preferentially selects intrinsically brighter objects and, by cutting out some of the distribution, narrows the observed distribution. As such, we note that the bias correction correctly increases the weight of low $\mu$ and high $\sigma$ parametrisations, as those models would be subject to the most sample bias. We also note that not only does the best fit location for each parameter shift, but the shape of the posterior surface itself is significantly modified.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces.}
	\label{fig:imperfect}
\end{figure}

@@

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous correction from Section \ref{sec:imperfect}. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$ (perhaps we believe it is approximately $20$) we estimate that the average contribution from $\beta y$ is approximately $20\beta = 4$ (which is close to the correct value of $6$), and from this we add in a small adjustment to the analytic correction from Section \ref{sec:imperfect}:
\begin{align}
w_{\rm approx} = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right].
\end{align}
This gives our likelihood parts as
\begin{align}
\mathcal{L}_{i1} &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}{w_{\rm approx} } \\
\mathcal{L}_{i2} &= \frac{ w_{\rm approx}  }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
Evaluating $\mathcal{L}_{i2}$ using Monte Carlo integration of $n$ samples, we have for a single observation that
\begin{align}
\mathcal{L}_{i2} = \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)} \label{eq:mc}.
\end{align}
We now wish to move to a set of observations. As $\mathcal{L}_{i2}$ is independent of the specific observation $i$, this is a simple step:
\begin{align}
\mathcal{L}_1 &= w_{\rm approx}^{-N} \prod_{i=1}^N \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y) \\
\mathcal{L}_{2} &= \left(  \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)}\right)^N
\end{align}
$\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample given by fitting $\mathcal{L}_{1}$. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either --- each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable.

Taking our concrete example stated previously, we end up with a corrected posterior surface as shown in Figure \ref{fig:real}. It is important to note that the point of maximum likelihood is biased by roughly the same amount in the biased and approximate posterior surfaces in Figure \ref{fig:real}. However, as the approximately correctly posterior is broader than the biased surface, it has far more samples in the region of parameter space mapped by the fully corrected posterior. This is the entire purpose of the approximate correction --- to maximise the number of samples in the correct region of parameter space, so that we can importance sample our chains efficiently.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{example/real.pdf}
	\end{center}
	\caption{A systematic test of our more complicated model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The likelihood $\mathcal{L}_1$ was evaluated, and reweighted using Monte Carlo integration of a hundred thousand possible events as per $\mathcal{L}_2$. The truncated data with no correction is shown as `Biased' in dotted red, the `Approximate` only correction ($\mathcal{L}_1$) shown in dashed green, and the final reweighted chain shown in solid black as `Corrected'. A large number of samples of $\mathcal{L}_1$ had to be generated to ensure sufficient sampling of our `Corrected' posterior.}
	\label{fig:real}
\end{figure}

>>>>

 In this section we present sample selections of increasing complexity. Each sample selection is accompanied by an illustrative example inspired by Type Ia supernova cosmology, where we characterize the properties of a standard or standardizable candle. Python code implementations of all examples, including plot generation, can be found at the Github repository \verb|SampleSelection|\footnote{\url{https://github.com/samreay/SampleSelection}}.

@@

We thus begin by considering an ideal situation where the sample is complete.  All events are included in the sample such that $P(S | {\rm data}, \theta)=1$.  Trivially this expression is independent of our data and model parameters $\theta$, and our likelihood from Equation \eqref{eq:main} reduces down to Equation \eqref{eq:like}. As a concrete example of this case, let us consider a model for a population of objects whose brightnesses form a normal distribution with average $\mu$ and standard deviation $\sigma$. Let us also assume that our experiment produces data $x_i$, which are independent measurements of unique objects' brightnesses (with negligible measurement uncertainty), which we can formally state as
\begin{align}
\vec{x} \sim \mathcal{N}(\mu,\sigma).
\end{align}
If, having collected our observations $\vec{x}$, we wanted to constrain $\mu$ and $\sigma$, this would be a simple task of modelling the posterior surface. We simply wish to map the surface
\begin{align}
P(\theta | {\rm data}) \propto P({\rm data} | \theta) P(\theta),
\end{align}
where our model parameters $\theta = \lbrace \mu, \sigma \rbrace$ and our data is given by $\vec{x}$.
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto P(\vec{x} | \mu, \sigma) P(\mu, \sigma)
\end{align}
With uniform priors, $P(\mu,\sigma) = {\rm constant}$, and can be absorbed into the constant of proportionality. Expanding our observation vector to take into account all observations and not just one, the posterior surface is given by
\begin{align}
P(\mu,\sigma| \vec{x}) &\propto \prod_{i=1}^N \mathcal{N}(x_i | \mu, \sigma), \label{eq:prod}
\end{align}
where $\mathcal{N}(x_i|\mu,\sigma)$ represents a the probability of drawing sample $x_i$ from the Normal distribution with mean $\mu$ and standard deviation $\sigma$. Generating a hundred data points with $\mu=100,\ \sigma=10$, we can recover our input parameters easily, as shown in Figure \ref{fig:perfect}.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{fig_0_perfect.pdf}
	\end{center}
	\caption{A systematic test of our perfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points. Any systematic offset in our model would be revealed by a shift in the stacked results away from the true parameter values. This plot can be generated by executing  \mbox{\texttt{0\_perfect.py}}.}
	\label{fig:perfect}
\end{figure}

@@

With this sample selection, all events satisfy $x_i > \alpha$, giving $P(S|x,\theta) = \mathcal{H}(x - \alpha)$, where $\mathcal{H}$ is the Heaviside step function
\begin{align}
\mathcal{H}(y)\equiv \begin{cases}
1 \quad {\rm if }\  y \ge 0 \\
0 \quad {\rm otherwise.}
\end{cases}
\end{align} 
We assign an arbitrary value $\alpha=85$ for convenience. If we do not take this truncation into account, we will recover biased parameter estimates. However, we can correct for this truncation using Equation \eqref{eq:main}, as the integral in the denominator has an analytic solution. Having successfully observed $x_i$, it follows that $x_i > \alpha$ and so $P(S|x_i,\theta) = 1$. To substitute in our normal model,
\begin{align}
\mathcal{L}_i &= \frac{P(S|x_i,\theta) P(x_i|\theta)}{\int P(S | D, \theta) P(D|\theta)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{-\infty}^\infty \mathcal{H}(D - \alpha) \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\int_{\alpha}^\infty \mathcal{N}(D|\mu, \sigma)\, dD} \\
&= \frac{ \mathcal{N}(x_i|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
where in the last line we have evaluated the integral in the case $\mu > \alpha$. Note that this is for a single observation, and so for a set of independent observations we need to introduce the product found in Equation \eqref{eq:prod}. 
\begin{align}
\mathcal{L} &= \prod_{i=1}^N \mathcal{L}_i =  \prod_{i=1}^N \frac{ \mathcal{N}(x|\mu, \sigma)}{\frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right]}, 
\end{align}
However, as our selection efficiency correction is independent of our observations $\vec{x}$, we can take it outside the product.
\begin{align}
\mathcal{L} &= 2 \left( {\rm erfc}\left[ \frac{\alpha - \mu}{\sqrt{2}\sigma} \right] \right)^{-N}\prod_{i=1}^N  \mathcal{N}(x|\mu, \sigma),
\end{align}
We can add this correction to our model, and note that we now recover unbiased parameter estimates. This is demonstrated in Figure \ref{fig:imperfect}, which shows the posterior surfaces for when you take sample selection into account and when you do not. The selection bias preferentially selects intrinsically brighter objects and, by cutting out some of the distribution, narrows the observed distribution. As such, we note that the bias correction correctly increases the weight of low $\mu$ and high $\sigma$ parametrisations, as those models would be subject to the most sample bias. We also note that not only does the best fit location for each parameter shift, but the shape of the posterior surface itself is significantly modified.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{fig_1_imperfect.pdf}
	\end{center}
	\caption{A systematic test of our imperfect model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The bias shown in the red `Biased' contour can be corrected to via the techniques shown in Section \ref{sec:imperfect} to recover unbiased surfaces. This plot can be generated by executing  \mbox{\texttt{1\_imperfect.py}}.}
	\label{fig:imperfect}
\end{figure}

@@

In our example, if $\beta \ll 1$, such that the majority of selection effect is encapsulated by $x$ and not $y$, our approximate correction can take the form found in the previous correction from Section \ref{sec:imperfect}. Having true values of $\mu = 100$, $\sigma = 10$, $\mu_y = 30$, $\sigma_y = 5$, $\alpha=92$ and a known $\beta = 0.2$, we can give a concrete example. Assuming some prior, imperfect knowledge of $\mu_y$ (perhaps we believe it is approximately $20$) we estimate that the average contribution from $\beta y$ is approximately $20\beta = 4$ (which is close to the correct value of $6$), and from this we add in a small adjustment to the analytic correction from Section \ref{sec:imperfect}:
\begin{align}
w_{\rm approx} = \frac{1}{2} {\rm erfc}\left[ \frac{\alpha - \mu - 4}{\sqrt{2}\sigma} \right].
\end{align}
This gives our likelihood parts as
\begin{align}
\mathcal{L}_{i1} &= \frac{ \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y)}{w_{\rm approx} } \\
\mathcal{L}_{i2} &= \frac{ w_{\rm approx}  }{\iint_{-\infty}^\infty \mathcal{H}(x + \beta y - \alpha) \mathcal{N}(X|\mu, \sigma) \mathcal{N}(Y|\mu_y, \sigma_y)\, dX dY}.
\end{align}
Evaluating $\mathcal{L}_{i2}$ using Monte Carlo integration of $n$ samples, we have for a single observation that
\begin{align}
\mathcal{L}_{i2} = \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)} \label{eq:mc}.
\end{align}
We now wish to move to a set of observations. As $\mathcal{L}_{i2}$ is independent of the specific observation $i$, this is a simple step:
\begin{align}
\mathcal{L}_1 &= w_{\rm approx}^{-N} \prod_{i=1}^N \mathcal{N}(x_i|\mu, \sigma) \mathcal{N}(y_i|\mu_y, \sigma_y) \\
\mathcal{L}_{2} &= \left(  \frac{ w_{\rm approx}  n }{\sum_{j=1}^{n} \mathcal{H}(X_j + \beta Y_j - \alpha) \mathcal{N}(X_j|\mu, \sigma) \mathcal{N}(Y_j|\mu_y, \sigma_y)}\right)^N
\end{align}
$\mathcal{L}_1$ can thus be fitted with a traditional sampler without numeric difficulty or slowdown, and $\mathcal{L}_2$ allows us to calculate the weight of each sample given by fitting $\mathcal{L}_{1}$. We are effectively importance sampling our likelihood evaluations. The computational benefits of this should not be understated either --- each sample in our chains can be reweighted independently, providing a task that is trivially parallelisable.

Taking our concrete example stated previously, we end up with a corrected posterior surface as shown in Figure \ref{fig:real}. It is important to note that the point of maximum likelihood is biased by roughly the same amount in the biased and approximate posterior surfaces in Figure \ref{fig:real}. However, as the approximately correctly posterior is broader than the biased surface, it has far more samples in the region of parameter space mapped by the fully corrected posterior. This is the entire purpose of the approximate correction --- to maximise the number of samples in the correct region of parameter space, so that we can importance sample our chains efficiently.
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{fig_2_real.pdf}
	\end{center}
	\caption{A systematic test of our more complicated model, done by stacking the output chains from fitting 100 independent realisations of our 100 data points, subject to our thresholding. The likelihood $\mathcal{L}_1$ was evaluated, and reweighted using Monte Carlo integration of a hundred thousand possible events as per $\mathcal{L}_2$. The truncated data with no correction is shown as `Biased' in dotted red, the `Approximate` only correction ($\mathcal{L}_1$) shown in dashed green, and the final reweighted chain shown in solid black as `Corrected'. A large number of samples of $\mathcal{L}_1$ had to be generated to ensure sufficient sampling of our `Corrected' posterior.  This plot can be generated by executing \mbox{\texttt{2\_real.py}}.}
	\label{fig:real}
\end{figure}
